---
title: "detected missing protein unique peptides analysis"
output:
  pdf_document: default
  html_document: default
date: "2023-04-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/cgu3/Documents/missing-protein-project/data/')
# run following line if you need to run code in console
# knitr::opts_knit$set(root.dir only change working directory of code chucks not code directly running in console
# setwd('/Users/cgu3/Library/CloudStorage/OneDrive-InsideMDAnderson/proteomics/missing protien project manuscript/missing-protein-project-data')
# getwd()
```


```{r}
library(tidyverse)
# if (!require("protools", quietly = TRUE))
#    devtools::install_github("https://github.com/FDUguchunhui/protools")
# library(protools)
```



# Import peptides of identified missing proteins
```{r}
MP_final_peptide <- readxl::read_xlsx('Supplementary file 2 identified missing protein details.xlsx', sheet = 2)
head(MP_final_peptide)
```

# Import uniqueness checking
```{r}
uniqueness_checking <- readxl::read_xlsx('Supplementary file 2 identified missing protein details.xlsx', sheet = 3)
head(uniqueness_checking)
```


Processing data before computing the number of unique peptides for each
identified missing proteins
```{r}
MP_final_peptide$Source <- str_extract(MP_final_peptide$Source, '^IPAS[0-9]+')
MP_final_peptide$unique <- ifelse((MP_final_peptide$unique =='.' | MP_final_peptide$unique == 'N'), 0 , 1)
```

# Import peptides of 177 identified missing proteins
```{r}
# MP_204_products <- readxl::read_xlsx('Supplementary file 2 identified missing protein details.xlsx', sheet = 1)
MP_177_products <- readxl::read_xlsx('Supplementary file 2 identified missing protein details.xlsx', sheet = 1)
MP_177_products$IPAS <- paste0('IPAS', str_extract(MP_177_products$IPAS, '(?<=IP).+(?=_)'))

```

# Calculate the number of unique peptides for each detected MPs (and identified MPs)
```{r}
unique_peptide_summary <- MP_final_peptide %>% filter(unique == 1) %>% group_by(protein.Accession, Source) %>% summarise(n=n(), unique_peptide=paste0(peptide.seq, collapse = '+')) %>% arrange(protein.Accession, desc(n), .by_group = FALSE) %>% ungroup()
```


# Get Summary of number of unique peptides for each of the 177 MPs
```{r}
unique_peptide_summary <- MP_177_products %>% left_join(unique_peptide_summary, by=c('accession' = 'protein.Accession', 'IPAS' = 'Source')) %>% replace_na(list(n=0))
# write_csv(unique_peptide_summary, 'missing_protein_unique_peptides.csv')
```


# plot unique peptide frequency for MP  with/without RNA expression
```{r}
unique_peptide_summary <-  unique_peptide_summary %>% mutate(count_discrete=ifelse(n >= 30, '30+', n))
unique_peptide_summary$count_discrete <- factor(unique_peptide_summary$count_discrete, levels=c(as.character(0:29), '30+'))

unique_peptide_summary %>% 
  ggplot(aes(x=count_discrete)) +
  geom_bar(position = 'identity', fill='red') +
theme_bw() +
  xlab('Matched unique peptide counts') +
  ylab('Frequency') +
  ggtitle('Matched unique peptide frequency of identified 177 missing proteins products') +
  scale_x_discrete(breaks=c('1', seq(0, 29, 5), '30+'), drop=FALSE)
  # scale_y_continuous(limits=c(0, 30))
```
```{r}
unique_peptide_summary %>% summary()
```

```{r}
unique_peptide_summary %>% filter(n >= 2)
```



# plot unique peptide frequency for MP with RNA expression
```{r}
unique_peptide_summary_88 <- unique_peptide_summary %>% filter(TPM > 0)
unique_peptide_summary_88 <-  unique_peptide_summary_88 %>% mutate(count_discrete=ifelse(n >= 30, '30+', n))
unique_peptide_summary_88$count_discrete <- factor(unique_peptide_summary_88$count_discrete, levels=c(as.character(0:29), '30+'))

unique_peptide_summary_88 %>% 
  ggplot(aes(x=count_discrete)) +
  geom_bar(position = 'identity', fill='blue') +
theme_bw() +
  xlab('Matched unique peptide counts') +
  ylab('Frequency') +
  ggtitle('Matched unique peptide frequency of identified 88 missing proteins products 
          with mRNA expression') +
  scale_x_discrete(breaks=c('1', seq(0, 29, 5), '30+'), drop=FALSE) 
```

```{r}
unique_peptide_summary_88 %>% summary()
```


```{r}
unique_peptide_summary_88 %>% filter(n >= 2)
```

