---
title: "cancer-specificity-analysis"
output:
  pdf_document: default
  html_document: default
date: "2023-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Volumes/hanash/Hanash_Lab_GPFS/Chunhui/missing protien project manuscript/missing-protein-project-data')
# run following line if you need to run code in console
# knitr::opts_knit$set(root.dir only change working directory of code chucks not code directly running in console
# setwd('/Users/cgu3/Library/CloudStorage/OneDrive-InsideMDAnderson/proteomics/missing protien project manuscript/missing-protein-project-data')
# getwd()
```


```{r}
library(tidyverse)
# Some function Reuqires the in-house package "protools"
if (!require("protools", quietly = TRUE))
   devtools::install_github("https://github.com/FDUguchunhui/protools")
library('protools')
```


# import data
nasf_primary_cell_2.csv: all primary cell data filtered with at least 2 spectral 
count and then normalized to NSAF
cell-line-raw-data/nasf_tce_2.csv: all cell line data  filtered with at least 2 spectral 
count and then normalized to NSAF
IPAS_annotation.csv: contain disease and subtype information for above samples
```{r}
# misspelling in the original file name 
primary_cell_nsaf <- read.csv('primary-cell/nasf_primary_cell_2.csv', row.names = 1)
primary_cell_nsaf <- as.matrix(primary_cell_nsaf)
tce_nsaf <- read.csv('cell-line-raw-data/nasf_tce_2.csv', row.names = 1)
tce_nsaf <- as.matrix(tce_nsaf)
# read annotation file
IPAS_annotation <- read_csv('support-data/IPAS_annotation.csv')
```





# create a helper function for creating annotation file for the expression matrix
```{r}
create_annoataion <- function(data, annotation_table) {
  annotation_tbl <- annotation_table %>% select(ipas, disease, subtype) %>% filter(ipas %in% colnames(data))
  annotation_df <- data.frame(disease=annotation_tbl$disease, subtype=annotation_tbl$subtype)
  rownames(annotation_df) <- annotation_tbl$ipas
  return(annotation_df)
}

primary_cell_annotation <- create_annoataion(primary_cell_nsaf, annotation_table = IPAS_annotation)
primary_cell_annotation$type <- paste(primary_cell_annotation$disease, primary_cell_annotation$subtype)
primary_cell_annotation <- primary_cell_annotation['type']
colnames(primary_cell_annotation) <- 'disease'
table(primary_cell_annotation$disease)
```

# create annotation dataframe for cell-line TCE data
```{r}
tce_annotation <- create_annoataion(tce_nsaf, annotation_table = IPAS_annotation)
tce_annotation$type <- paste(tce_annotation$disease, tce_annotation$subtype)
tce_annotation <- tce_annotation[, c('disease', 'type')]
colnames(tce_annotation) <- c('cancer', 'subtype')
# check distribution of each cancer and subtype
table(tce_annotation$cancer)
table(tce_annotation$subtype)
```


# import missing protein
```{r}
missing_proteins <- readxl::read_xlsx('Supplementary file 2 identified missing protein details.xlsx', sheet = 1)
unique_67_accession  <- missing_proteins %>% filter(TPM > 0) %>% select(accession) %>% pull() %>% unique()
```



# create a varaible to denote cancer-testis for the identified MPs
```{r}
cancer_testis_gene <- c('WFDC11','SPATA21','C19orf67','CCNYL2','TRAV9-2','TRAV6','TMEM105','NCBP2L','TMC3','PCDHAC1','A3GALT2','OR2L2','CTAGE15','RGPD1','SLAMF9','ASB18','TRAV34','PPIAL4G','RFPL4AL1','ANKUB1','GRXCR2','PRSS48','R3HDML')
missing_proteins_annotation <- missing_proteins %>% filter(TPM > 0) %>%  select(accession, gene_symbol) %>% unique()
missing_proteins_annotation <- missing_proteins_annotation %>% mutate(cancer_testis=if_else(gene_symbol %in% cancer_testis_gene, 1, 0))
missing_proteins_annotation <- missing_proteins_annotation %>% arrange(cancer_testis)
```


# extract the IPAS to keep 
For primary cell, only use following cancers:
"Leukemia ALL", "Leukemia AML", "Leukemia MDS", "Ovarian Ascites", "Gastric Ascites"

For cell-line TCE, only use the following cancers/subtypes
"Breast HER2", "Breast LuminalA/B", "Breast TNBC", "Gastric Ascites",
"Gastric Adeno", "Leukemia AML", "LungAdeno Mesenchymal", "LungAdeno Epithelial",
"LungAdeno NA", "Ovarian Adeno ", "Pancreatic ExocrineAdeno", "SCLC Neurendocrine"

The reason for not using all cancer/subtype is because some subtypes only have
few samples, and including them will make heatmap plot color hard to read
```{r}
extract_sample_id <- function(x, disease_col, disease) {
  rownames(x$annotation[x$annotation[[disease_col]] %in% disease, ,drop=FALSE])
}

# diseases_keep <- c('Colon', 'Glioma', 'Leukemia', 'Lymphoblast', 'Melanoma', 'Neuroblastoma', 'Prostate', 'Ovarian')
diseases_keep <- c("Leukemia ALL", "Leukemia AML", "Leukemia MDS", "Ovarian Ascites", "Gastric Ascites")
SpC_list_primary_cell <- SpC_List(primary_cell_nsaf, annotation = primary_cell_annotation, proteins_filter = unique_67_accession)
replicates_for_keep <- extract_sample_id(SpC_list_primary_cell, disease_col = 'disease', disease = diseases_keep)


# diseases_keep <- c('Colon', 'Glioma', 'Leukemia', 'Lymphoblast', 'Melanoma', 'Neuroblastoma', 'Prostate', 'Ovarian')
diseases_keep_tce <- c("Breast HER2", "Breast LuminalA/B", "Breast TNBC", "Gastric Ascites", "Gastric Adeno", "Leukemia AML", "LungAdeno Mesenchymal", "LungAdeno Epithelial", "LungAdeno NA", "Ovarian Adeno ", "Pancreatic ExocrineAdeno", "SCLC Neurendocrine")
SpC_list_tce <- SpC_List(tce_nsaf, annotation = tce_annotation, proteins_filter = unique_67_accession)
replicates_for_keep_tce <- extract_sample_id(SpC_list_tce, disease_col = 'subtype', disease = diseases_keep_tce)
```



# extract the IPAS to keep for each cancer/subtype
This information is later used to reorder samples in matrix for better plotting heatmap
```{r}
IPAS <- mapply(FUN=extract_sample_id, disease=c("Leukemia ALL", "Leukemia AML", "Leukemia MDS", "Ovarian Ascites", "Gastric Ascites"), MoreArgs = list(x=SpC_list_primary_cell, disease_col='disease'))
# SpC_lists <- mapply(FUN=SpC_List, replicates_keep=IPAS, MoreArgs = list(df=primary_cell_nasf_gene_symbol, annotation=primary_cell_annotation, proteins_filter=MP_204_gene_symbol), SIMPLIFY = FALSE)
IPAS_tce <- mapply(FUN=extract_sample_id, disease= c("Breast HER2", "Breast LuminalA/B", "Breast TNBC", "Gastric Ascites", "Gastric Adeno", "Leukemia AML", "LungAdeno Mesenchymal", "LungAdeno Epithelial", "LungAdeno NA", "Ovarian Adeno", "Pancreatic ExocrineAdeno", "SCLC Neurendocrine"), MoreArgs = list(x=SpC_list_tce, disease_col='subtype'))
```


# reorder samples to make samples of the same cancer/subtype next to each other
```{r}
primary_cell_nasf <- primary_cell_nsaf[, unlist(IPAS)]
tce_nsaf <- tce_nsaf[, unlist(IPAS_tce)]
```


# Create the final data object used for both primaru cell and cell-line TCE
SpC_List is a convenient customized object for simplifying analysis pipeline
See ?SpC_List for more details
```{r}
all_cancer_primary_cell <- SpC_List(df=primary_cell_nasf, annotation=primary_cell_annotation, proteins_filter = unique_67_accession)
all_cancer_tce <- SpC_List(df=tce_nsaf, annotation=tce_annotation, proteins_filter = unique_67_accession)
```



# create cancer-testis row annotation for primary cell
```{r}
all_cancer_primary_cell$matrix <- all_cancer_primary_cell$matrix[missing_proteins_annotation$accession, ]

row_annotation <- missing_proteins_annotation[, c("accession", "cancer_testis")]
row_annotation <- data.frame(`cancer testis`=row_annotation$cancer_testis, row.names = row_annotation$accession)
row_annotation$cancer.testis <- as.factor(row_annotation$cancer.testis)
```


```{r, fig.width=12, fig.height=8}
plot_heat_map(all_cancer_primary_cell,
         annotation_row = row_annotation,
         cluster_rows = T,
         cluster_cols = F,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer primary cell data',
         show_colnames = F,)
```


```{r, fig.width=12, fig.height=8}
plot_heat_map(all_cancer_primary_cell,
         annotation_row = row_annotation,
         cluster_rows = F,
         cluster_cols = F,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer primary cell data',
         show_colnames = F,)
```





# reorder to make caqncer-testis perotein to bottom part
```{r}
# find those MPs that were empty in data
empty_MPs <- missing_proteins_annotation$accession[which(!(missing_proteins_annotation$accession %in% rownames(all_cancer_tce$matrix)))]
additional_matrix <- matrix(0, nrow = length(empty_MPs), ncol = ncol(all_cancer_tce$matrix))
rownames(additional_matrix) <- empty_MPs
all_cancer_tce$matrix <- rbind(all_cancer_tce$matrix, additional_matrix)
all_cancer_tce$matrix <- all_cancer_tce$matrix[missing_proteins_annotation$accession, ]
```


```{r, fig.width=12, fig.height=8}
row_annotation <- missing_proteins_annotation[, c("accession", "cancer_testis")]
row_annotation <- data.frame(`cancer testis`=row_annotation$cancer_testis, row.names = row_annotation$accession)
row_annotation$cancer.testis <- as.factor(row_annotation$cancer.testis)
```


```{r, fig.width=12, fig.height=8}
(cancer_specificity_cell_line_row_clustered <- plot_heat_map(all_cancer_tce,
         annotation_row = row_annotation,
         cluster_rows = T,
         cluster_cols = F,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer cell-line data',
         show_colnames = F))

# ggsave('~/Downloads/cancer_specificity_primary_cell_row_clustered.png', cancer_specificity_primary_cell_row_clustered,
#        width = 1020, height=735, units = 'px', dpi=80)
```


```{r, fig.width=12, fig.height=8}
(cancer_specificity_cell_line <- plot_heat_map(all_cancer_tce,
         annotation_row = row_annotation,
         cluster_rows = F,
         cluster_cols = F,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer cell-line data',
         show_colnames = F))
```



```{r, fig.width=12, fig.height=8}
(cancer_specificity_cell_line_col_clustered <- plot_heat_map(all_cancer_tce,
         annotation_row = row_annotation,
         cluster_rows = T,
         cluster_cols = F,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer cell-line data',
         show_colnames = F))
```


```{r, fig.width=12, fig.height=8}
(cancer_specificity_cell_line_col_clustered <- plot_heat_map(all_cancer_tce,
         annotation_row = row_annotation,
         cluster_rows = T,
         cluster_cols = T,
         show_rownames = T,
         fontsize_row=7,
              main='Cancer-specificity for the identied 67 missing proteins in cancer cell-line data',
         show_colnames = F))
```


